---
title: "Ancova satu jalur menggunakan R"
author: "Gede Saindra Santyadiputra"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
<a href="https://doi.org/10.5281/zenodo.10212024" target="_blank">![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10212024.svg)</a>

## Load the library

```{r warning=FALSE}
library(conflicted)
library(rstatix)
library(gridExtra)
library(dplyr)
library(ggpubr)
library(rstatix)
library(broom)
library(readr)
library(tidyverse)
```

## Load the dataset

```{r}
df58=read_csv("data58.csv")
df58
```

## Summary statistics

### Summary statistics for dependent variable posttest

```{r}
df58 %>% group_by(VLEs) %>%  get_summary_stats(posttest, type="common")
```

### Summary statistics for covariate pretest

```{r}
df58 %>% group_by(VLEs) %>%  get_summary_stats(pretest, type="common")
```

### Data visualization

```{r}
p1 <- ggplot(df58, aes(pretest, posttest, colour = VLEs)) + geom_point(size = 3) + theme(legend.position="top")
p2 <- ggplot(df58, aes(x = VLEs, y = posttest, col = VLEs)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2) + theme(legend.position="top")
p3 <- ggplot(df58, aes(x = VLEs, y = pretest, fill = VLEs)) + geom_boxplot(outlier.shape = NA) + geom_jitter(width = 0.2) + theme(legend.position="top")
grid.arrange(p1, p2, p3, ncol=3)
```

## Model fit

### Fit the model, the covariate goes first

```{r}
model <- lm(posttest ~ pretest + VLEs, data = df58)
```

### Inspect the model diagnostic metrics and remove details

```{r}
model.metrics <- augment(model) %>%
  select(-.hat, -.sigma, -.fitted)
head(model.metrics, 3)
```

## Assumption tests

### Outliers above 3.00

```{r}
model.metrics %>% 
  dplyr::filter(abs(.std.resid) > 3) %>%
  as.data.frame()
```

### Outliers above 2.50

```{r}
model.metrics %>% 
  dplyr::filter(abs(.std.resid) > 2.5) %>%
  as.data.frame()
```

### Linearity

```{r}
ggscatter(
  df58, x = "pretest", y = "posttest",
  color = "VLEs", add = "reg.line"
)+
  stat_regline_equation(
    aes(label =  paste(..eq.label.., ..rr.label.., sep = "~~~~"), color = VLEs)
  )
```

### Homogeneity of regression slopes

```{r}
df58 %>% anova_test(posttest ~ VLEs*pretest)
```

### Normality of residuals

```{r}
shapiro_test(model.metrics$.resid)
```

### Homogeneity of variances

```{r}
model.metrics %>% levene_test(.resid ~ VLEs)
```

## ANCOVA

### ANCOVA table

```{r}
anova_test(data = df58, formula = posttest ~ pretest + VLEs, type = 3, detailed = TRUE)
```

### Estimated Marginal Means (EMM)

```{r}
adj_means <- emmeans_test(data = df58, formula = posttest ~ VLEs, covariate = pretest)
get_emmeans(adj_means)
```

### Post-hoc test: bonferroni

```{r}
emmeans_test(data = df58, formula = posttest ~ VLEs, covariate = pretest, p.adjust.method = "bonferroni")
```

## Referensi

1. <a href="https://www.datanovia.com/en/lessons/Ancova-in-r/" target="_blank">Ancova in R</a>
2. <a href="https://www.reneshbedre.com/blog/Ancova.html" target="_blank">Ancova using R and Python</a>

## Sitasi dokumen ini

`Santyadiputra, G. S., Purnomo, & Juniantari, M. (2023, November 28). Ancova satu jalur menggunakan R. RPubs. https://doi.org/10.5281/zenodo.10212024`

<a href="https://doi.org/10.5281/zenodo.10212024" target="_blank">![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.10212024.svg)</a>